Make any stupid plot you want

### 2021-02-24 draft/sections/intro.tex
pushed mpl down to vtk, reworked gog again

\subsection{Tools}
\label{sec:intro_data_tools}
The library driving this rearchitecture, Matplotlib, aims to be flexible enough that developers can build domain specific libraries on top of it. To preserve this flexibility, Matplotlib enforces minimal constraints on the types of data the user is allowed to input and supports very many types of graphics, from primitive marks to computationally complex visualizations. Instead of enforcing constraints at the API level, Matplotlib carries implicit assumptions about data continuity in how each function interfaces with the input data. This has lead to poor API consistency and brittle code as every visual algorithm has a very different point of view on how the data is structured.

A commonly cited alternative is the family of tools built on top of Wilkenson's Grammar of Graphics (GoG) \cite{wilkinsonGrammarGraphics2005}, including ggplot\cite{wickhamGgplot2ElegantGraphics2016a}, protovis\cite{bostockProtoviz2009} and D3 \cite{bostockDataDrivenDocuments2011}, vega\cite{satyanarayanDeclarativeInteractionDesign2014} and altair\cite{vanderplasAltairInteractiveStatistical2018}. The GoG family of libraries are very strongly built on the assumption that data is a relational table, which restricts the native set of visualization types the libraries support. Since the data space and graphic space is very well defined in this grammar, it lends itself to a declarative interface \cite{heerDeclarative2010} that allows end users to describe the visualization they are trying to create. Grammar oriented approaches describe how to compose visual elements into a graphical design \cite{wongsuphasawatNavigatingWideWorld2021}, while we are proposing a framework for building those elements. An example of this distinction is that the GoG grammar includes computation and aggregation of the table as part of the grammar, while we propose that most computations are specific to domains and only try to describe them when they are specifically part of the visual encoding - for example mapping data to a color. Disentangling the computation from the visual transforms allows us to determine whether the visualization library needs to handle them or if they can be more efficiently computed by the data container.

A different class of user facing tools are those that support images, such as ImageJ\cite{schneiderNIHImageImageJ2012}. These tools mostly have some support for visualizing non image components of a complex data set, but mostly in service to the image being visualized. These tools are ill suited for general purpose library developers as the architecture is to build plugins into the existing system \cite{WritingPlugins} rather than domain specific tools that are built on top. Even the digital humanities oriented macro ImagePlot\cite{studiesCulturevisImageplot2021}, which supports some non-image aggregate reporting charts, is still built around image data as the primary input.

There are also visualization tools designed around scientifically complex data that support explicit description of the data, such as vtk\cite{hanwellVisualizationToolkitVTK2015,geveci2012vtk} and its derivatives such as MayaVi\cite{ramachandranMayaVi2011} and extensions such as ParaView\cite{ahrens2005paraview} and the infoviz themed Titan\cite{brianwylieUnifiedToolkitInformation2009}. These libraries, generally speaking, are architectured more as graphics libraries than visualization libraries, meaning they lack a clear distinction between the graphic construction and the rendering of that graphic. This is very close to the existing Matplotlib architecture, where the data, visual encoding, and rendering are jumbled in ways that make it hard to figure out what the code is doing. We are proposing a functional framework instead in large part to clearly encapsulate these separate responsibilities of a visualization tool. In turn, this should make it easier to reuse components in the building of new tools.

>>>>

\subsection{Tools}
\label{sec:intro_data_tools}
One extensive family of relational table based libraries are those based on Wilkenson's Grammar of Graphics (GoG) \cite{wilkinsonGrammarGraphics2005}, including ggplot\cite{wickhamGgplot2ElegantGraphics2016a}, protovis\cite{bostockProtoviz2009} and D3 \cite{bostockDataDrivenDocuments2011}, vega\cite{satyanarayanDeclarativeInteractionDesign2014} and altair\cite{vanderplasAltairInteractiveStatistical2018}. The restriction to tables in turn restricts the native design space to visualizations suited to tables. Since the data space and graphic space is very well defined in this grammar, it lends itself to a declarative interface \cite{heerDeclarative2010}. This grammar oriented approach allows users to describe how to compose visual elements into a graphical design \cite{wongsuphasawatNavigatingWideWorld2021}, while we are proposing a framework for building those elements. An example of this distinction is that the GoG grammar includes computation and aggregation of the table as part of the grammar, while we propose that most computations are specific to domains and only try to describe them when they are specifically part of the visual encoding - for example mapping data to a color. Disentangling the computation from the visual transforms allows us to determine whether the visualization library needs to handle them or if they can be more efficiently computed by the data container.

A different class of user facing tools are those that support images, such as ImageJ\cite{schneiderNIHImageImageJ2012} or Napari\cite{nicholas_sofroniew_2021_4533308}. These tools often have some support for visualizing non image components of a complex data set, but mostly in service to the image being visualized. These tools are ill suited for general purpose libraries that need to support data other than images because the architecture is oriented towards building plugins into the existing system \cite{WritingPlugins} where the image is the core data structure. Even the digital humanities oriented ImageJ macro ImagePlot\cite{studiesCulturevisImageplot2021}, which supports some non-image aggregate reporting charts, is still built around image data as the primary input.

There are also visualization tools where there is no single core structure, and instead internally carry around many different representations of data. Matplotlib, has this structure, as does VTK \cite{hanwellVisualizationToolkitVTK2015,geveci2012vtk} and its derivatives such as MayaVi\cite{ramachandranMayaVi2011} and extensions such as ParaView\cite{ahrens2005paraview} and the infoviz themed Titan\cite{brianwylieUnifiedToolkitInformation2009}. Where GoG and ImageJ type libraries have very consistent APIs for their visualization tools because the data structure is the same, the APIs for visualizations in VTK and Matplotlib are significantly dependent on the structure of the data it expects. This in turn means that every new type of visualization must carry implicit assumptions about data structure in how it interfaces with the input data. This has lead to poor API consistency and brittle code as every visualization type has a very different point of view on how the data is structured. This API choice particularly breaks down when the same dataset is fed into visualizations with different assumptions about structure or into a dashboard consisting of different types of visualization\cite{a.sarikayaWhatWeTalk2019,fewDashboardConfusionRevisited2007} because there is no consistent way to update the data and therefore no consistent way of guaranteeing that the views stay in sync. Our model is a structure dependent formalism, but then also provides a core representation of that structure that is abstract enough to provide a common interface for many different types of visualization.

### 2021-02-24 draft/sections/math.tex
pushed mpl down to vtk, reworked gog again

\subsubsection{Sheaf}
\label{sec:artist_sheaf_stalk}
As part of the definition of local triviality, there is an open neighborhood $U\subset \dbase$ for every $\dbasepoint \in \dbase$. We can define the inclusion map $\iota: U \rightarrow \dbase$  which pulls \dtotal over $U$ 
\begin{equation}
    \label{eq:sheaf}
    \begin{tikzcd}
        \iota^*\dtotal \arrow[d, "\pi"'] \arrow[r, "\iota^*", hook]             & \dtotal \arrow[d, "\pi"']                  \\
        U \arrow[r, "\iota", hook] \arrow[u, "\iota^{*}\dsection"', bend right] & \dbase \arrow[u, "\dsection"', bend right]
    \end{tikzcd}
\end{equation}
such that the pulled back $\iota^*\dsection$ only contains records over $U \subset \dbase$. Restriction maps on the sheaf define how to glue together these localized sections $\iota^*\dsection$ into larger sections\cite{ghristElementaryAppliedTopology2014,ghristHomologicalAlgebraData2018} over the subset of \dbase\ requested by the visualization.

The sheaf facilitates visualizations such as sliding windows\cite{crouchDynamicGraphsSlidingwindow2013,chuTimeSeriesSegmentation1995} streaming data, or navigation such as pan and zoom\cite{NekrasovskiEvaluationPanZoom2006}.;

@@

For example, a scatter plot of red circles is the output of one artist, a scatter plot of green squares the output of another, as shown in figure~\ref{fig:artist_equivalance}. These two artists are equivalent since their only difference is in the literal visual encodings (color, shape). Shape and color could also be defined in \vmark\, but the position must come from the fiber $\vfiber=(xpos,ypos)$ since fundementally a scatter plot is the plotting of one position against another\cite{friendlyBriefHistoryData2008}. We also use this criteria to identify derivative types, for example the bubble chart\cite{tufteVisualDisplayQuantitative2001} is a type of scatter where by definition the glyph size is mapped from the data. The criteria for equivalence class membership serves as the basis for evaluating invariance\cite{kindlmann2014algebraic}.

\subsection{Making the fiber bundle computable}
\label{sec:triangulization}
\begin{figure}[H]
    \includegraphics{figures/math/simplex.png}
    \caption{Simplices can encode the connectivity of the data, from fully disconnected (0 simplex) records to all records are connected to at least 3 others}
    \label{fig:triangle_simplex}
\end{figure}
%%big data
One way of expressing the connectivity of records in a dataset is to implement \dbase\ as a simplacial complex, which is a set of simplicies such as those shown in figure~\ref{fig:triangle_simplex}. The advantage of triangulization is that it is general enough to work for more complex topology based visualization methods \cite{heineSurveyTopologybasedMethods2016} while also providing a consistent interface of vertices, edges, and faces for \vindex\ to map into. When triangulated, the simplices encode the continuity in the data

\begin{table}[H]
    \begin{center}
        \begin{tabular}{|l|l|l|}\hline
        \textbf{simplex} & \textbf{continuity} & \textbf{\dsection}   \\ \hline
        vertex  & discrete   & \dsection(\dbasepoint)                  \\ \hline
        edge    & 1D         & \dsection(\dbasepoint,\, \dx)        \\ \hline
        face    & 2D         & \dsection(\dbasepoint,\, \dx,\, \dy)\\ \hline
        \end{tabular}
        \caption{}
        \label{tab:triangulization}
    \end{center}
\end{table}

such that each section is bound to a simplex $\dbasepoint \in \dbase$. As shown in table~\ref{tab:triangulization}, in a 1D continuous spaces each \dsection\ lies distance \dx\ along edge \dbasepoint, while in a 2D continuous space each \dsection\ lies at coordinate \dx,\dy on the face \dbasepoint. This is directly analogous to indexing to express connectivity in N-D arrays, while also natively supporting graphs and trees as they are simplacial complices of nodes and edges. Path connected components are then sections where edges or faces meet.

In general simplacial complexes, in these prototypes we implement graphs.

\end{document}

>>>>

\subsubsection{Sheaf}
\label{sec:artist_sheaf_stalk}
The restriction maps of a sheaf describe how local \dsections can be glued into larger sections \cite{ghristElementaryAppliedTopology2014,ghristHomologicalAlgebraData2018}. As part of the definition of local triviality, there is an open neighborhood $U\subset \dbase$ for every $\dbasepoint \in \dbase$. We can define the inclusion map $\iota: U \rightarrow \dbase$  which pulls \dtotal over $U$ 
\begin{equation}
    \label{eq:sheaf}
    \begin{tikzcd}
        \iota^*\dtotal \arrow[d, "\pi"'] \arrow[r, "\iota^*", hook]             & \dtotal \arrow[d, "\pi"']                  \\
        U \arrow[r, "\iota", hook] \arrow[u, "\iota^{*}\dsection"', bend right] & \dbase \arrow[u, "\dsection"', bend right]
    \end{tikzcd}
\end{equation}
such that the pulled back $\iota^*\dsection$ only contains records over $U \subset \dbase$. By gluing $\iota^*\dsection$ together, the sheaf is putting a continuous structure on local sections which allows for defining a section over a subset in \dbase. That section over subset \dbase\ maps to the graphic generated by \vartist\ for visualizations such as sliding windows\cite{crouchDynamicGraphsSlidingwindow2013,chuTimeSeriesSegmentation1995} streaming data, or navigation techniques such as pan and zoom\cite{NekrasovskiEvaluationPanZoom2006}.

@@

For example, a scatter plot of red circles is the output of one artist, a scatter plot of green squares the output of another, as shown in figure~\ref{fig:artist_equivalance}. These two artists are equivalent since their only difference is in the literal visual encodings (color, shape). Shape and color could also be defined in \vmark\, but the position must come from the fiber $\vfiber=(xpos,ypos)$ since fundementally a scatter plot is the plotting of one position against another\cite{friendlyBriefHistoryData2008}. We also use this criteria to identify derivative types, for example the bubble chart\cite{tufteVisualDisplayQuantitative2001} is a type of scatter where by definition the glyph size is mapped from the data. The criteria for equivalence class membership serves as the basis for evaluating invariance\cite{kindlmann2014algebraic}. 
\end{document}
